Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-24 paper/paper.tex
pass through discussion section. questions.

\section{Discussion}
\label{sec:discuss}
\yl{I'm not sure how strong of a statement we can make about our proof
technique. We've already demonstrated how to address type-level
dependency when defining a logical relation through a simple
example. Claiming that our proof structure is better seems quite
ambitious, but I think there's a middle point where we claim that
adding moderate features like typed reduction doesn't instantly make
our code size expand all the way from 1000 to 20,000 without saying
the other developments are just verbose for no good reason}
The short consistency proof we
present achieves the goal of demonstrating the technique of proof by
logical relation for dependently typed languages. However, what
remains unanswered is what makes our development significantly shorter
compared to others. Are we proving simpler results, or is our proof
technique genuinely more efficient?
% focus on the following question: why is our proof,
% even with its extension to the existence $\beta\eta$-normal form, so much
% shorter than the other mechanized results from \citet{decagda,
% nbeincoq, martin-lof-a-la-coq}?

First, the metatheoretic property we prove is indeed simpler.
Unlike developments that mechanize the correctness of a
type-directed conversion algorithm, we only show the existence of
normal form for open and closed terms and state our properties in
terms of the untyped small-step reduction relation. This removes a lot
of scaffolding related to the specification of the algorithm and the
extra proof obligation that the algorithm is sound and complete with
respect to the declarative specification of the type system.
As a result, our logical relation, unlike the ones from \citet{decagda,
  nbeincoq, martin-lof-a-la-coq, anand2014towards}, maps from types to
predicates rather than relations. The need for a relational model is
directly related to the metatheoretic property one wants to
prove. \citet{anand2014towards} requires a relational model to capture
the extensionality of their type system, whereas \citet{decagda,
  martin-lof-a-la-coq, nbeincoq} uses a relational model to derive the
injectivity of $\Pi$ types and justify the validity of
$\eta$-conversion among other properties.
Since \lang{} uses an untyped conversion rule, type conversion can be
done through the \emph{normalize-and-compare}
strategy described in \citet{pierce2004advanced}. The decidability of
normalize-and-compare is implied by the existence of
$\beta\eta$-normal form, which follows from our logical
predicate.

In terms of our proof technique, we close over our semantically valid types
and terms in \rref{I-Par, I-Bool} using the non-deterministic parallel reduction relation,
which relates to full reduction, whereas \citet{decagda,nbeincoq} all
refer to the deterministic weak head reduction. The use of a
non-deterministic reduction strategy comes at the cost that we need
confluence to prove the functionality of our logical relation, but
immediately gives us a semantic justification of the conversion rule
(Lemma~\ref{lemma:logrelcoherence}).
% In particular, we obtain the confluence
% result of reduction at a very early stage before we even define our
% logical relation.

With untyped conversion,
we sidestep the relational, Kripke-style logical relation found in
other mechanized proofs. However, our early dependence on confluence
before the fundamental theorem is established can be alarming.
In a system with type-directed reduction,
confluence is not immediately available since it depends on
depends on $\Pi$-injectivity, which is usually only proven after the
fundamental theorem.
Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
problem that allow us to recover confluence property independently
from the logical relation. \citet{siles2012pure} generalize the
notion of Type Parallel One Step Reduction from \citet{adams2006pure}
to syntactically prove $\Pi$-injectivity for arbitrary Pure Tyep
Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
equational theory, thus allowing subject reduction to be proven
independently from confluence. Therefore, we believe that even in a
system where type-directed reduction is required (e.g. a system with
the unit $\eta$-law) in the logical relation, the proof can still be carried
out in a structure similar to the one we have presented.

Finally, despite our earlier claim that our metatheoretic properties
are not as strong as some of the related work, we can strengthen our
results through syntactic means. For example, the
normalize-and-compare strategy does not induce an efficient algorithm
for type conversion, like the ones from \citet{decagda} and
\citet{martin-lof-a-la-coq}. However, with the standard syntactic
techniques from \citet{takahashi-parallel-reduction,
factorization-essentially}, we can prove that leftmost-outermost
reduction is a normalizing reduction strategy, from which we can
separately show the correctness a more efficient algorithm that
reduces the two terms to weak head normal form before recursively
comparing their subcomponents. We believe factoring such properties
out of a logical relation is valuable, as it helps us identify the
part of our proof that requires extra strength from the metatheory.

>>>>

\section{Discussion}
\label{sec:discuss}
\yl{I'm not sure how strong of a statement we can make about our proof
  technique. We've already demonstrated how to address type-level dependency
  when defining a logical relation through a simple example. Claiming that our
  proof structure is better seems quite ambitious, but I think there's a
  middle point where we claim that adding moderate features like typed
  reduction doesn't instantly make our code size expand all the way from 1000
  to 20,000 without saying the other developments are just verbose for no good
  reason} The short consistency proof we present achieves the goal of
demonstrating the technique of proof by logical relation for dependently typed
languages. However, what remains unanswered is what makes our development
significantly shorter compared to others. Are we proving simpler results, for
smaller languages, are we making more use of automation, or is our proof
technique genuinely more efficient?
% focus on the following question: why is our proof,
% even with its extension to the existence $\beta\eta$-normal form, so much
% shorter than the other mechanized results from \citet{decagda,
% nbeincoq, martin-lof-a-la-coq}?

First, the metatheoretic properties that we prove are indeed simpler.  Unlike
developments that mechanize the correctness of a type-directed conversion
algorithm, we only show the existence of normal forms for open and closed
terms and state our properties in terms of an untyped reduction relation. This
avoids a lot of the scaffolding related to the specification of the algorithm
and the proof obligations that the algorithm is sound and complete with
respect to the declarative specification of the type system.
As a result, our logical relation, unlike the ones from \citet{decagda,
  nbeincoq, martin-lof-a-la-coq, anand2014towards}, maps from types to
predicates rather than relations. \scw{How does this follow? Why do we need unary relations where they need binary relations?} The need for a relational model is
directly related to the metatheoretic property one wants to
prove. \citet{anand2014towards} requires a relational model to capture
the extensionality of their type system, whereas \citet{decagda,
  martin-lof-a-la-coq, nbeincoq} uses a relational model to derive the
injectivity of $\Pi$ types and justify the validity of
$\eta$-conversion among other properties.
Since \lang{} uses an untyped conversion rule, type conversion can be
done through the \emph{normalize-and-compare}
strategy described in \citet{pierce2004advanced}. The decidability of
normalize-and-compare is implied by the existence of
$\beta\eta$-normal form, which follows from our logical
predicate.

\scw{I'm getting confused by this paragraph. Does this reorganization sense:
  Our language is simpler than Nuprl, because it doesn't have extensional
  equality. It is simpler than Agda, because it doesn't have type-directed
  equality. Both of these cases require the definition of a binary logical
  relation, that defines a notion of semantic equality between terms. This
  relation justifies the injectivity of $\Pi$ types and justify the validity
  of $\eta$-conversion among other properties.}  \scw{Furthermore our proof is
  also simpler because we don't need prove the correctness of the NBE
  algorithm, which is used to show the decidability of Agda's type-directed
  equivalence. Therefore, we don't need to define this algorithm and show that
  it is sound and complete with respect to the type-directed
  equality. Instead, to show the decidability of our untyped equivalence, we
  need only show that terms have $\beta\eta$ normal forms. }

In terms of our proof technique, we close over our semantically valid types
and terms in \rref{I-Par, I-Bool} using the non-deterministic parallel
reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak
head reduction relation. Our use of a non-deterministic reduction strategy
means that we need confluence to prove the functionality of our logical
relation. However, the benefit is that it immediately gives us a semantic
justification of the conversion rule (Lemma~\ref{lemma:logrelcoherence}).
% In particular, we obtain the confluence
% result of reduction at a very early stage before we even define our
% logical relation.

With untyped conversion,
we sidestep the relational, Kripke-style logical relation found in
other mechanized proofs. \scw{Need to define Kripke-style. Also the other 
proofs need Kripke style because they are defining typed relations, not untyped 
relations. }
However, our early dependence on confluence
before the fundamental theorem is established can be alarming.
In a system with type-directed reduction,
confluence is not immediately available because it 
depends on $\Pi$-injectivity, which is usually only proven after the
fundamental theorem.\scw{confluence depends on Pi injectivity? I thought 
it was only needed for subject reduction}
Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
problem that allow us to recover confluence property independently
from the logical relation. \citet{siles2012pure} generalize the
notion of Type Parallel One Step Reduction from \citet{adams2006pure}
to syntactically prove $\Pi$-injectivity for arbitrary Pure Type
Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
equational theory, thus allowing subject reduction to be proven
independently from confluence. Therefore, we believe that even in a
system where type-directed reduction is required (e.g. a system with
the unit $\eta$-law) in the logical relation, the proof can still be carried
out in a structure similar to the one we have presented. \scw{Not sure that 
I understand this paragraph}

Finally, despite our earlier claim that our metatheoretic properties
are not as strong as some of the related work, we can strengthen our
results through syntactic means. For example, the
normalize-and-compare strategy does not induce an efficient algorithm
for type conversion, like the ones from \citet{decagda} and
\citet{martin-lof-a-la-coq}. However, with the standard syntactic
techniques from \citet{takahashi-parallel-reduction,
factorization-essentially}, we can prove that leftmost-outermost
reduction is a normalizing reduction strategy, from which we can
separately show the correctness a more efficient algorithm that
reduces the two terms to weak head normal form before recursively
comparing their subcomponents. We believe factoring such properties
out of a logical relation is valuable, as it helps us identify the
part of our proof that requires extra strength from the metatheory.
\scw{Why don't we just use leftmost-outermost reduction in the first place? 
Do we even need nondeterministic parallel reduction?}
